---
title: "Predictors of Trust in AI"
subtitle: "Summary of Preliminary Results"
author: "Scott Claessens"
date: "`{r} Sys.Date()`"
format:
  html:
    html-table-processing: none
    embed-resources: true
toc: true
execute:
  echo: false
  warning: false
  error: false
---

```{r load-packages}
library(targets)
library(tidyverse)
```

This document summarises the preliminary results of our study on the predictors
of trust in artificial intelligence systems.

We recruited 800 participants (N = 690 after exclusions for low Captcha scores
and failed attention check) from Prolific. In the study, participants answered
questions about both "general AI" and a random subset of specific AI systems (5 
out of 20). In each case, we asked participants how much they trusted the system
and various other questions, including perceived reliability, competence,
genuine nature, ethicality, autonomy, potential for good, potential for harm,
interpretability, explainability, anthropomorphism, and predictability. We also
collected data on whether participants had heard of and used the different AI
systems.

Data and code for the study can be found here:
<https://github.com/ScottClaessens/person-trust-AI>

## Reported usage of different AI systems

We found that some AI systems were more familiar to participants than others.

```{r}
#| label: fig-usage
#| fig-cap: "Proportion of participants that have heard of and used different AI systems"
#| out-width: "100%"
#| out-height: "66%"

tar_read(plot_usage)
```

## Reported trust in different AI systems

We ranked average ratings of trust for "general AI" and the different AI 
systems. The results show that participants trust some AI systems more than
others, with "general AI" sitting somewhere in the middle.

```{r}
#| label: fig-trust-ranking
#| fig-cap: "Self-reported trust in different AI systems"
#| out-width: "100%"
#| out-height: "80%"

tar_read(plot0_trust)
```

## Predictors of trust in general AI

For general AI, we looked at the correlation matrix for self-reported trust and
all other questions.

```{r}
#| label: fig-cor-general
#| fig-cap: "Correlation matrix for 'General AI'"
#| out-width: "100%"
#| out-height: "75%"

tar_read(plot_cor_General.AI)
```

We found that perceived reliability, competence, genuine nature, and ethicality
have the strongest positive correlations with trust. This makes sense, since 
previous work has suggested that trust in AI is a multidimensional concept that
is made up of both performance trust (reliability and competence) and moral 
trust (genuine nature and ethicality; Malle & Ullman, 2021). Trust is also 
strongly positively correlated with the belief that AI has the potential for 
good.

Other variables are also correlated with trust, but to a lesser degree.

## Predictors of trust in different AI systems

To assess how this pattern varies across different AI systems, we fitted a
series of multilevel models that allowed the effects of predictors to vary
across AI systems. The variance parameter for the varying slopes allows us to
determine how much the effect of each predictor on trust *varies* across 
different AI systems.

```{r}
#| label: fig-varying-slopes
#| fig-cap: "Variance in slopes across different AI systems"
#| out-width: "100%"
#| out-height: "73%"

tar_read(plot_slope_sds)
```

This plot suggests that, for example, the effect of interpretability on trust
does not vary much across different AI systems.

```{r}
#| label: fig-interpretability
#| fig-cap: "Effect of interpretability on trust across different AI systems"
#| out-width: "100%"
#| out-height: "83%"

tar_read(plot1_interpretability)
```

On the other hand, the effect of competence on trust varies much more across
different AI systems -- though the effect is still positive for all AI systems.

```{r}
#| label: fig-competence
#| fig-cap: "Effect of competence on trust across different AI systems"
#| out-width: "100%"
#| out-height: "83%"

tar_read(plot1_competent)
```

## Future directions

There is more that we can do with this dataset:

- **Which types of AI are most prototypical?** We pre-registered that we would
conduct a principal components analysis on all the variables and map the 
different AI types in the dimensional space to determine which are closest to
"general AI". In other words, which types of AI are people probably thinking 
about when surveys ask them about AI in general? Unfortunately for us, 
everything is highly correlated in this dataset, meaning that PCA just pulls out
a single dimension (which can be conceptualised as how "positive" or "negative"
participants feel).
- **How does trust vary across individuals?** We have data on demographics,
political ideology, and familiarity with AI, so we could potentially use these
as predictors of trust.
- **Are there any interaction effects?** We have looked at predictors 
separately. But the variables likely interact with each other. For example,
competence is probably a weaker predictor of trust for autonomous killer drones
because participants believe that this AI has the potential for harm -- they
don't want it to achieve that harm more competently!

Beyond these questions, we would like to scale this up across countries. Before
doing this, it would be good to get some direction on what is most interesting
to focus on going forward.

## References

Malle, B. F., & Ullman, D. (2021). A multidimensional conception and measure of 
human-robot trust. In *Trust in human-robot interaction* (pp. 3-25). Academic 
Press.
